---
title: "Environment Setup"
always_allow_html: true
---

This page loads all shared dependencies, authenticates to Google Earth Engine, and derives the study area boundary and hydrological base layers used throughout the analysis.

## Packages, GEE Authentication, and AOI

All computational pages source `_common.R`, which handles package loading, GEE authentication, helper function definitions, and AOI derivation from HydroSHEDS level-6 basins.

```{r}
#| label: setup
source("_common.R")
```

## Area of Interest

The workflow begins with the coordinates of a point location within the waterbody of interest. From there, we filter the surrounding basin boundary using the HydroSHEDS polygon dataset of global watersheds. In a first attempt, multiple levels may need querying to find the encompassing polygon.

```{r}
#| label: aoi-summary
print(paste("Level 6 basins containing point:", nrow(basin_l6_sf)))
print(paste("Level 8 sub-catchments:", nrow(basin_l8_sf)))
```

If a local watershed shapefile exists, overlay it against the GEE-derived boundary to assess agreement. With the AOI established, the level-6 basin becomes the primary clip boundary for all subsequent raster and vector queries.

```{r}
#| label: aoi-export
sf::st_write(aoi_sf, "./outputs/aoi_sf.shp", delete_dsn = TRUE)
```

```{r}
#| label: aoi-map

tmap_mode("view")
tm_shape(st_as_sf(aoi_sf)) + tm_borders(col = "red", lwd = 2) +
  tm_shape(basin_l8_sf) + tm_borders(col = "blue", lwd = 0.5, lty = "dashed") +
  tm_shape(st_as_sf(poi_sf)) + tm_symbols(fill = "green", size = 1) +
  tm_basemap("Esri.WorldImagery") +
  tm_scalebar(position = c("right", "bottom"))
```

## Hydrological Base Layers

We import elevation and processed hydrology data, including flow direction and accumulation, rivers, streams, and waterbody features, annual drainage estimates, monthly water surface elevation and channel width estimates, and monthly water reflectance values. We import these hydraulic layers from multiple sources for cartographic and data validation purposes: [MERIT Hydro](https://developers.google.com/earth-engine/datasets/catalog/MERIT_Hydro_v1_0_1) (v1.0.3), [HydroSHEDS](https://www.hydrosheds.org) (v1.1), and [JRC Global Surface Water](https://developers.google.com/earth-engine/datasets/catalog/JRC_GSW1_4_GlobalSurfaceWater) (v1.4). Hydrology in these landscapes requires cross-validating between multiple methods and sources, such as comparing HydroSHEDS stream networks with streams derived from MERIT's upstream drainage layer capped at 10 km^2^ catchment area.

```{r}
#| label: hydro-rasters

# HydroSHEDS: conditioned DEM (~90m)
dem              <- ee$Image("WWF/HydroSHEDS/03VFDEM")$clip(aoi_ee)
hydro_flowdir    <- ee$Image("WWF/HydroSHEDS/03DIR")$clip(aoi_ee)
hydro_flowacc    <- ee$Image("WWF/HydroSHEDS/03ACC")$clip(aoi_ee)
hydro_rivers     <- ee$FeatureCollection("WWF/HydroSHEDS/v1/FreeFlowingRivers")$filterBounds(aoi_ee)
hydro_lakes      <- ee$FeatureCollection("projects/sat-io/open-datasets/HydroLakes/lake_poly_v10")$filterBounds(aoi_ee)
hydro_rivers_sf  <- ee_to_sf(hydro_rivers)
hydro_lakes_sf   <- ee_to_sf(hydro_lakes)

# MERIT Hydro: flow direction, upstream drainage, water elevation, & channel width
merit_flowdir    <- ee$Image("MERIT/Hydro/v1_0_1")$select("dir")$clip(aoi_ee)
merit_drainup    <- ee$Image("MERIT/Hydro/v1_0_1")$select("upa")$clip(aoi_ee)
merit_elev       <- ee$Image("MERIT/Hydro/v1_0_1")$select("elv")$clip(aoi_ee)
merit_wth        <- ee$Image("MERIT/Hydro/v1_0_1")$select("wth")$clip(aoi_ee)
merit_rivers     <- merit_drainup$gt(10)

# JRC Global Surface Water
gsw_occurrence   <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("occurrence")$clip(aoi_ee)
gsw_seasonality  <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("seasonality")$clip(aoi_ee)
gsw_recurrence   <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("recurrence")$clip(aoi_ee)
gsw_transition   <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("transition")$clip(aoi_ee)
gsw_monthly      <- ee$ImageCollection("JRC/GSW1_4/MonthlyHistory")$filterBounds(aoi_ee)

print(paste("JRC monthly images:", gsw_monthly$size()$getInfo()))
print(paste("Number of lakes within watershed:", nrow(hydro_lakes_sf)))
```

```{r}
#| label: hydro-map

tmap_mode("view")
tm_shape(st_as_sf(aoi_sf)) + tm_borders(col = "red", lwd = 2) +
  tm_shape(hydro_rivers_sf) + tm_lines(col = "steelblue", lwd = 1.5) +
  tm_shape(hydro_lakes_sf) + tm_polygons(fill = "steelblue", lwd = 1) +
  tm_shape(st_as_sf(poi_sf)) + tm_symbols(fill = "green", size = 1) +
  tm_scalebar(position = c("right", "bottom")) +
  tm_basemap("Esri.WorldImagery")
```
