---
title: "Google Earth Engine setup via rgee"
subtitle: "Lake Chilwa Basin — run this first from RStudio"
author: "Murphy, S."
date: today
format:
  html:
    toc: true
    code-fold: false
engine: knitr
execute:
  eval: false   # set to true once your credentials are configured
---

## What this is

An R-first (rgee) setup for Earth Engine. rgee wraps the Python
`earthengine-api` through reticulate, so you control everything from R with
negligible speed overhead relative to raw reticulate. Run this once to confirm
credentials and dataset access before any analysis notebook.

## Credentials: read this first

Earth Engine does **not** authenticate with a `GOOGLE_API_KEY`. It uses an
OAuth user credential created by `ee_Authenticate()`, tied to a Google Cloud
project that has the Earth Engine API enabled. Your `GOOGLE_API_KEY` in
`.Renviron` is for other Google APIs and is ignored here. Because your usual EE
credentials are not present, you will run `ee_Authenticate()` once below to
create them.

You also need a project ID. Add `EE_PROJECT=your-project-id` to `.Renviron`
(the project must have the Earth Engine API enabled at
console.cloud.google.com). Restart R after editing `.Renviron`.

## One-time install

```{r install}
#| eval: false
install.packages("rgee")
# Build the repo-local .venv (identical to the manuscript's environment-setup
# chunk), so earthengine-api lives inside the project and is called via the
# reticulate <-> Quarto bridge on Render.
venv_dir <- here::here(".venv")
base_py  <- Sys.getenv("RETICULATE_PYTHON", unset = "/opt/local/bin/python")
reticulate::virtualenv_create(venv_dir, python = base_py)
reticulate::virtualenv_install(venv_dir, packages = c("earthengine-api", "numpy"))
reticulate::use_virtualenv(venv_dir, required = TRUE)
rgee::ee_check()             # verifies Python, earthengine-api, numpy in .venv
```

## Authenticate (once) and initialise

```{r auth-init}
library(rgee)
# First run only: opens a browser, creates and stores the EE credential.
# ee_Authenticate()

# Initialise against your project. drive = TRUE also authorises Google Drive so
# ee_*_to_drive() exports work; omit it if you will not export.
ee_Initialize(project = Sys.getenv("EE_PROJECT"), drive = TRUE)
ee_user_info()
```

## Smoke test: datasets over the AOI

```{r smoke}
library(sf)

poi <- ee$Geometry$Point(c(35.55, -15.35))
aoi <- ee$FeatureCollection("WWF/HydroSHEDS/v1/Basins/hybas_6")$
  filterBounds(poi)$geometry()

# rgee helper: bring an EE geometry back to R as sf (replaces custom ee_to_sf)
aoi_sf <- ee_as_sf(aoi)
print(st_area(aoi_sf))

# Sentinel-1 GRD (C-band SAR)
s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
  filterBounds(aoi)$
  filter(ee$Filter$eq("instrumentMode", "IW"))$
  filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VV"))$
  filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VH"))$
  filter(ee$Filter$eq("orbitProperties_pass", "DESCENDING"))
cat("Sentinel-1 scenes:", s1$size()$getInfo(), "\n")

# Landsat Collection 2 L2
for (id in c("LANDSAT/LT05/C02/T1_L2","LANDSAT/LE07/C02/T1_L2","LANDSAT/LC08/C02/T1_L2"))
  cat(id, ":", ee$ImageCollection(id)$filterBounds(aoi)$size()$getInfo(), "scenes\n")

# rgee interactive map (replaces the ee_tile_url + tmap tile hack)
gsw <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("occurrence")
Map$centerObject(aoi, zoom = 9)
Map$addLayer(gsw$clip(aoi), list(min = 0, max = 100, palette = c("white","blue")),
             "JRC surface-water occurrence")
```

If the scene counts print and the map renders, rgee is working and the
manuscript's `environment-setup` chunk will initialise the same way. Next:
`05.scripts/rgee-workflow-design.md` for the full R-first workflow and the
speed guidance.
