---
title: "GEE Watershed Data Sourcing: Lake Chilwa Basin"
author: "Murphy, S."
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "View Code"
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
engine: knitr
---

## Setup

Initialise GEE via reticulate, matching the main analysis environment.

```{r setup}
library(sf)
library(tmap)
library(dplyr)

reticulate::use_python("/opt/local/bin/python", required = TRUE)
ee <- reticulate::import("ee")
ee$Initialize(project = "murphys-deforisk")

# Helpers
ee_tile_url <- function(ee_image, vis_params) {
  ee_image$getMapId(vis_params)$tile_fetcher$url_format
}

ee_to_sf <- function(fc) {
  geojsonsf::geojson_sf(
    jsonlite::toJSON(fc$getInfo(), auto_unbox = TRUE)
  )
}
```

## Step 1: Select a Point

The entire workflow begins from a single coordinate pair placed within
the waterbody of interest. Everything downstream derives from this point.

```{r point}
# Lake Chilwa centre (lon, lat)
lon <- 35.55
lat <- -15.35

poi <- ee$Geometry$Point(c(lon, lat))
poi_sf <- st_sfc(st_point(c(lon, lat)), crs = 4326)
```

## Step 2: Identify the Containing Watershed

Query HydroSHEDS basins at multiple levels to find the watershed polygon
that contains the selected point. Level 6 returns the broadest drainage
basin; level 8 gives finer sub-catchments nested within it.

```{r watershed-query}
# Broad basin (level 6)
basin_l6 <- ee$FeatureCollection("WWF/HydroSHEDS/v1/Basins/hybas_6")$
  filterBounds(poi)
basin_l6_sf <- ee_to_sf(basin_l6)

# Sub-catchments (level 8)
basin_l8 <- ee$FeatureCollection("WWF/HydroSHEDS/v1/Basins/hybas_8")$
  filterBounds(poi)
basin_l8_sf <- ee_to_sf(basin_l8)

print(paste("Level 6 basins containing point:", nrow(basin_l6_sf)))
print(paste("Level 8 sub-catchments:", nrow(basin_l8_sf)))
```

## Step 3: Define the AOI from the Watershed

Use the level-6 basin as the primary AOI geometry. This becomes the
clip boundary for all subsequent raster and vector queries.

```{r aoi}
# Dissolve level-6 features into a single watershed boundary
aoi_sf <- st_union(basin_l6_sf)
aoi_ee <- ee$FeatureCollection(basin_l6)$geometry()

# Bounding box for map framing
aoi_bbox <- st_bbox(aoi_sf)
```

## Step 4: Visual Check

Confirm the point falls within the derived watershed and that the
boundary aligns with expected drainage divides.

```{r map-check}
tmap_mode("view")

tm_shape(st_as_sf(aoi_sf)) +
  tm_borders(col = "red", lwd = 2) +
tm_shape(basin_l8_sf) +
  tm_borders(col = "cyan", lwd = 0.5, lty = "dashed") +
tm_shape(st_as_sf(poi_sf)) +
  tm_dots(col = "yellow", size = 0.1) +
tm_basemap("Esri.WorldImagery") +
tm_scalebar(position = c("right", "bottom"))
```

## Step 5: Compare with Local Shapefile

If a local watershed shapefile exists, overlay it against the
GEE-derived boundary to assess agreement.

```{r compare-local}
local_aoi <- read_sf("../inputs/chilwa_watershed_4326.shp")

tmap_mode("view")

tm_shape(st_as_sf(aoi_sf)) +
  tm_borders(col = "red", lwd = 2) +
tm_shape(local_aoi) +
  tm_borders(col = "blue", lwd = 2, lty = "dashed") +
tm_shape(st_as_sf(poi_sf)) +
  tm_dots(col = "yellow", size = 0.1) +
tm_basemap("Esri.WorldImagery") +
tm_scalebar(position = c("right", "bottom"))
```

## Step 6: Clip Hydrological Rasters to Watershed

With the AOI established, source DEM and flow derivatives from
HydroSHEDS and MERIT Hydro, clipped to the watershed boundary.

```{r hydrorasters}
# HydroSHEDS: conditioned DEM (~90m)
hydrodem <- ee$Image("WWF/HydroSHEDS/03VFDEM")$clip(aoi_ee)

# HydroSHEDS: D8 flow direction
flowdir <- ee$Image("WWF/HydroSHEDS/03DIR")$clip(aoi_ee)

# HydroSHEDS: flow accumulation
flowacc <- ee$Image("WWF/HydroSHEDS/03ACC")$clip(aoi_ee)
```

```{r merit}
# MERIT Hydro: flow direction
merit_dir <- ee$Image("MERIT/Hydro/v1_0_1")$select("dir")$clip(aoi_ee)

# MERIT Hydro: upstream drainage area (km2)
merit_upa <- ee$Image("MERIT/Hydro/v1_0_1")$select("upa")$clip(aoi_ee)

# MERIT Hydro: water surface elevation
merit_elv <- ee$Image("MERIT/Hydro/v1_0_1")$select("elv")$clip(aoi_ee)

# MERIT Hydro: river channel width
merit_wth <- ee$Image("MERIT/Hydro/v1_0_1")$select("wth")$clip(aoi_ee)
```

## Step 7: Cartographic Hydrography

Rivers, streams, and lake polygons for base map context. These are
cartographic reference layers, not analytical inputs.

```{r rivers}
# HydroSHEDS free-flowing rivers (global vector, filtered to AOI)
rivers <- ee$FeatureCollection("WWF/HydroSHEDS/v1/FreeFlowingRivers")$
  filterBounds(aoi_ee)
rivers_sf <- ee_to_sf(rivers)
print(paste("River segments in AOI:", nrow(rivers_sf)))
```

```{r lakes}
# HydroLAKES: global lake polygons (community upload)
lakes <- ee$FeatureCollection(
  "projects/sat-io/open-datasets/HydroLakes/hydrolakes_polys")$
  filterBounds(aoi_ee)
lakes_sf <- ee_to_sf(lakes)
print(paste("Lakes in AOI:", nrow(lakes_sf)))
```

```{r streams}
# MERIT Hydro: derive a stream network from upstream drainage area
# Threshold at 10 km2 catchment area to extract visible streams
merit_upa_clip <- ee$Image("MERIT/Hydro/v1_0_1")$
  select("upa")$clip(aoi_ee)
streams_raster <- merit_upa_clip$gt(10)
```

```{r map-hydro}
tmap_mode("view")

tm_shape(st_as_sf(aoi_sf)) +
  tm_borders(col = "red", lwd = 2) +
tm_shape(lakes_sf) +
  tm_fill(col = "steelblue", alpha = 0.5) +
  tm_borders(col = "steelblue", lwd = 1) +
tm_shape(rivers_sf) +
  tm_lines(col = "dodgerblue", lwd = 1.5) +
tm_shape(st_as_sf(poi_sf)) +
  tm_dots(col = "yellow", size = 0.1) +
tm_basemap("Esri.WorldImagery") +
tm_scalebar(position = c("right", "bottom"))
```

## Step 8: Source Surface Water History

JRC Global Surface Water provides Landsat-derived water occurrence,
seasonality, and monthly presence maps spanning 1984 to 2021. Clip to
the watershed AOI.

```{r jrc-gsw}
gsw <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$clip(aoi_ee)

gsw_occurrence   <- gsw$select("occurrence")
gsw_seasonality  <- gsw$select("seasonality")
gsw_recurrence   <- gsw$select("recurrence")
gsw_transition   <- gsw$select("transition")

# Monthly water history within the watershed
gsw_monthly <- ee$ImageCollection("JRC/GSW1_4/MonthlyHistory")$
  filterBounds(aoi_ee)
print(paste("JRC monthly images:", gsw_monthly$size()$getInfo()))
```

## Step 8: Export

Write GEE-sourced vectors to the staging directory for review.

```{r export}
st_write(basin_l6_sf,
  "../_staging/chilwa_hydrobasins_l6.gpkg", delete_dsn = TRUE)
st_write(basin_l8_sf,
  "../_staging/chilwa_hydrobasins_l8.gpkg", delete_dsn = TRUE)
st_write(rivers_sf,
  "../_staging/chilwa_rivers.gpkg", delete_dsn = TRUE)
st_write(lakes_sf,
  "../_staging/chilwa_lakes.gpkg", delete_dsn = TRUE)
```

## Summary

The workflow proceeds in a single chain from point to watershed to data:

1. Place a coordinate in the waterbody of interest.
2. Query HydroSHEDS basins to retrieve the containing watershed at two
   nesting levels.
3. Dissolve the broad basin into a single AOI geometry.
4. Visually confirm the boundary and compare with any local shapefiles.
5. Clip HydroSHEDS DEM derivatives and MERIT Hydro layers to the AOI.
6. Source rivers, streams, and lake polygons for cartographic context.
7. Clip JRC surface water history to the AOI.
8. Export vectors to `_staging/` for review.
