---
title: "Sentinel-1 SAR Processing: Lake Chilwa Basin"
author: "Murphy, S."
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "View Code"
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
engine: knitr
---

## Setup

```{r setup}
library(sf)
library(tmap)
library(dplyr)

reticulate::use_python("/opt/local/bin/python", required = TRUE)
ee <- reticulate::import("ee")
ee$Initialize(project = "murphys-deforisk")

ee_tile_url <- function(ee_image, vis_params) {
  ee_image$getMapId(vis_params)$tile_fetcher$url_format
}

ee_to_sf <- function(fc) {
  geojsonsf::geojson_sf(
    jsonlite::toJSON(fc$getInfo(), auto_unbox = TRUE)
  )
}
```

## AOI

Derive the watershed AOI from a point, following the workflow in
`gee-watershed-sourcing.qmd`.

```{r aoi}
poi <- ee$Geometry$Point(c(35.55, -15.35))

basin_l6 <- ee$FeatureCollection("WWF/HydroSHEDS/v1/Basins/hybas_6")$
  filterBounds(poi)
aoi_ee <- basin_l6$geometry()
aoi_sf <- ee_to_sf(basin_l6) |> st_union()
```

## Step 1: Query Sentinel-1 GRD Archive

GEE hosts Sentinel-1 Ground Range Detected (GRD) products with
radiometric calibration and terrain correction already applied via the
Sentinel-1 Toolbox. The archive begins in October 2014. We filter for
IW mode, VV and VH polarisations, and descending orbit to maintain
consistent viewing geometry.

```{r s1-query}
s1 <- ee$ImageCollection("COPERNICUS/S1_GRD")$
  filterBounds(aoi_ee)$
  filter(ee$Filter$eq("instrumentMode", "IW"))$
  filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VV"))$
  filter(ee$Filter$listContains("transmitterReceiverPolarisation", "VH"))$
  filter(ee$Filter$eq("orbitProperties_pass", "DESCENDING"))$
  select(c("VV", "VH", "angle"))

n_scenes <- s1$size()$getInfo()
print(paste("Total Sentinel-1 scenes:", n_scenes))

# Date range
dates <- s1$reduceColumns(
  ee$Reducer$minMax(), list("system:time_start"))$getInfo()
print(paste("First scene:", as.Date(as.POSIXct(dates[[1]] / 1000, origin = "1970-01-01"))))
print(paste("Last scene:", as.Date(as.POSIXct(dates[[2]] / 1000, origin = "1970-01-01"))))
```

## Step 2: Speckle Filtering

GEE does not apply speckle filtering by default. We implement a focal
mean approximation. For refined Lee Sigma filtering, export to SNAP.
The focal approach is sufficient for time series compositing where
multi-temporal averaging further suppresses speckle.

```{r speckle-filter}
apply_speckle_filter <- function(image) {
  # Focal mean with 7x7 kernel (approximates boxcar filter)
  vv_filtered <- image$select("VV")$
    focal_mean(radius = 3.5, kernelType = "square", units = "pixels")$
    rename("VV_filtered")
  vh_filtered <- image$select("VH")$
    focal_mean(radius = 3.5, kernelType = "square", units = "pixels")$
    rename("VH_filtered")

  image$addBands(vv_filtered)$addBands(vh_filtered)$
    copyProperties(image, list("system:time_start"))
}

s1_filtered <- s1$map(apply_speckle_filter)
```

## Step 3: Backscatter Ratio and Difference Bands

VV/VH ratio and difference help distinguish open water from flooded
vegetation. Open water shows low VV and low VH. Flooded vegetation
shows moderate VV but elevated VH due to double-bounce scattering.

```{r ratio-bands}
add_ratio_bands <- function(image) {
  vv <- image$select("VV_filtered")
  vh <- image$select("VH_filtered")

  # Ratio (in dB domain, ratio = difference)
  ratio <- vv$subtract(vh)$rename("VV_VH_ratio")

  # Normalised difference
  nd <- vv$subtract(vh)$divide(vv$add(vh))$rename("VV_VH_nd")

  image$addBands(ratio)$addBands(nd)$
    copyProperties(image, list("system:time_start"))
}

s1_processed <- s1_filtered$map(add_ratio_bands)
```

## Step 4: Water Detection Thresholding

Calm water surfaces yield low backscatter in both VV and VH channels,
typically below -15 dB for VV. We use a percentile-based adaptive
threshold: compute the VV distribution within the AOI on a reference
scene and take the value that separates the lowest backscatter mode
(water) from land. A fixed fallback of -15 dB is available if needed.

```{r water-threshold}
# Reference scene for threshold calibration
sample_image <- ee$Image(s1_processed$first())$select("VV_filtered")

# Compute percentile-based threshold within AOI
# The 15th percentile of VV typically captures the water/land break
# in scenes with substantial water coverage like Chilwa
percentiles <- sample_image$reduceRegion(
  reducer = ee$Reducer$percentile(c(15L, 50L)),
  geometry = aoi_ee,
  scale = 30L,
  maxPixels = as.integer(1e9)
)$getInfo()

threshold_vv <- percentiles$VV_filtered_p15
print(paste("Adaptive VV threshold (p15):", round(threshold_vv, 2), "dB"))
print(paste("Median VV:", round(percentiles$VV_filtered_p50, 2), "dB"))

# Use adaptive threshold; fall back to -15 dB if percentile is unreasonable
if (is.null(threshold_vv) || threshold_vv > -10 || threshold_vv < -25) {
  threshold_vv <- -15
  print("Using fixed threshold: -15 dB")
}

classify_water <- function(image) {
  water <- image$select("VV_filtered")$lt(threshold_vv)$rename("water")
  image$addBands(water)$
    copyProperties(image, list("system:time_start"))
}

s1_water <- s1_processed$map(classify_water)
```

## Step 5: Monthly Composites

Aggregate to monthly composites to reduce data volume and suppress
residual speckle through temporal averaging.

```{r monthly-composites}
# Build year-month pairs in R, then iterate on the GEE side
year_month_pairs <- expand.grid(year = 2015L:2024L, month = 1L:12L)

composite_list <- lapply(seq_len(nrow(year_month_pairs)), function(i) {
  y <- as.integer(year_month_pairs$year[i])
  m <- as.integer(year_month_pairs$month[i])

  start <- ee$Date$fromYMD(y, m, 1L)
  end   <- start$advance(1L, "month")

  monthly <- s1_water$filterDate(start, end)
  n <- monthly$size()$getInfo()

  if (n == 0L) return(NULL)

  monthly$mean()$
    set("system:time_start", start$millis())$
    set("year", y)$
    set("month", m)$
    set("n_scenes", n)
})

# Drop empty months
composite_list <- Filter(Negate(is.null), composite_list)

s1_monthly <- ee$ImageCollection$fromImages(composite_list)
print(paste("Monthly composites:", length(composite_list)))
```

## Step 6: Time Series Extraction

Extract mean backscatter and water fraction within the AOI for each
monthly composite to build the SAR time series.

```{r time-series}
extract_stats <- function(image) {
  stats <- image$select(c("VV_filtered", "VH_filtered", "water"))$
    reduceRegion(
      reducer = ee$Reducer$mean(),
      geometry = aoi_ee,
      scale = 30L,
      maxPixels = as.integer(1e9)
    )
  ee$Feature(NULL, stats)$
    set("date", image$date()$format("YYYY-MM-dd"))$
    set("year", image$get("year"))$
    set("month", image$get("month"))$
    set("n_scenes", image$get("n_scenes"))
}

s1_ts <- s1_monthly$map(extract_stats)
s1_ts_info <- s1_ts$getInfo()

# Convert to data frame
s1_df <- do.call(rbind, lapply(s1_ts_info$features, function(f) {
  data.frame(
    date        = f$properties$date,
    year        = f$properties$year,
    month       = f$properties$month,
    n_scenes    = f$properties$n_scenes,
    VV_mean     = f$properties$VV_filtered,
    VH_mean     = f$properties$VH_filtered,
    water_frac  = f$properties$water,
    stringsAsFactors = FALSE
  )
}))

s1_df$date <- as.Date(s1_df$date)
print(head(s1_df))
```

## Step 7: Visual Check

Map a wet-season and dry-season composite side by side.

```{r map-composites}
# Select a wet month (January) and dry month (August) from same year
wet <- s1_monthly$
  filter(ee$Filter$eq("month", 1L))$
  filter(ee$Filter$eq("year", 2020L))$first()

dry <- s1_monthly$
  filter(ee$Filter$eq("month", 8L))$
  filter(ee$Filter$eq("year", 2020L))$first()

vis_sar <- list(bands = list("VV_filtered"), min = -25, max = 0,
  palette = c("black", "white"))
vis_water <- list(bands = list("water"), min = 0, max = 1,
  palette = c("white", "blue"))

# Get tile URLs for wet and dry composites
wet_sar_url   <- ee_tile_url(wet$clip(aoi_ee), vis_sar)
dry_sar_url   <- ee_tile_url(dry$clip(aoi_ee), vis_sar)
wet_water_url <- ee_tile_url(wet$clip(aoi_ee), vis_water)
dry_water_url <- ee_tile_url(dry$clip(aoi_ee), vis_water)

tmap_mode("view")

# Wet season: VV backscatter + water mask
map_wet <- tm_shape(st_as_sf(aoi_sf)) +
  tm_borders(col = "red", lwd = 2) +
  tm_basemap("Esri.WorldImagery") +
  tm_tiles(wet_sar_url, group = "Wet VV (Jan 2020)") +
  tm_tiles(wet_water_url, group = "Wet water mask (Jan 2020)") +
  tm_scalebar(position = c("right", "bottom"))

# Dry season: VV backscatter + water mask
map_dry <- tm_shape(st_as_sf(aoi_sf)) +
  tm_borders(col = "red", lwd = 2) +
  tm_basemap("Esri.WorldImagery") +
  tm_tiles(dry_sar_url, group = "Dry VV (Aug 2020)") +
  tm_tiles(dry_water_url, group = "Dry water mask (Aug 2020)") +
  tm_scalebar(position = c("right", "bottom"))

# Side by side
tmap_arrange(map_wet, map_dry, ncol = 2)
```

## Step 8: Export

Write the time series data frame and flag composites for potential
raster export.

```{r export}
write.csv(s1_df, "../_staging/s1_monthly_timeseries.csv", row.names = FALSE)

# To export a composite as GeoTIFF (uncomment and run as needed):
# task <- ee$batch$Export$image$toDrive(
#   image = wet$select(c("VV_filtered", "VH_filtered", "water"))$clip(aoi_ee),
#   description = "S1_wet_2020_01",
#   folder = "GEE_exports",
#   region = aoi_ee,
#   scale = 10L,
#   maxPixels = as.integer(1e9)
# )
# task$start()
```

## Notes

GEE Sentinel-1 GRD products arrive with radiometric calibration (sigma0)
and Range-Doppler terrain correction already applied. The processing
chain here adds speckle filtering (focal mean, 7x7), polarimetric ratio
bands, Otsu-based water thresholding, and monthly compositing. For
interferometric analysis or refined Lee Sigma filtering, export raw
scenes to SNAP.
