---
title: "Landsat Time Series Processing: Lake Chilwa Basin (1994-2015)"
author: "Murphy, S."
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "View Code"
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
engine: knitr
---

## Setup

```{r setup}
library(sf)
library(tmap)
library(dplyr)
library(ggplot2)

reticulate::use_python("/opt/local/bin/python", required = TRUE)
ee <- reticulate::import("ee")
ee$Initialize(project = "murphys-deforisk")

ee_to_sf <- function(fc) {
  geojsonsf::geojson_sf(
    jsonlite::toJSON(fc$getInfo(), auto_unbox = TRUE)
  )
}
```

## AOI

```{r aoi}
poi <- ee$Geometry$Point(c(35.55, -15.35))

basin_l6 <- ee$FeatureCollection("WWF/HydroSHEDS/v1/Basins/hybas_6")$
  filterBounds(poi)
aoi_ee <- basin_l6$geometry()
aoi_sf <- ee_to_sf(basin_l6) |> st_union()
```

## Step 1: Harmonise Landsat 5, 7, and 8 Surface Reflectance

Landsat Collection 2 Level-2 surface reflectance products span the
study period (1994-2015) across three sensors. Band names differ between
sensors and must be harmonised to a common naming convention before
compositing.

```{r band-harmonise}
# Landsat 5 TM: 1984-2012
rename_l5 <- function(image) {
  image$select(
    c("SR_B1", "SR_B2", "SR_B3", "SR_B4", "SR_B5", "SR_B7", "ST_B6", "QA_PIXEL"),
    c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2", "Thermal", "QA_PIXEL")
  )$multiply(0.0000275)$add(-0.2)$
    copyProperties(image, list("system:time_start", "system:index"))
}

# Landsat 7 ETM+: 1999-present
rename_l7 <- function(image) {
  image$select(
    c("SR_B1", "SR_B2", "SR_B3", "SR_B4", "SR_B5", "SR_B7", "ST_B6", "QA_PIXEL"),
    c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2", "Thermal", "QA_PIXEL")
  )$multiply(0.0000275)$add(-0.2)$
    copyProperties(image, list("system:time_start", "system:index"))
}

# Landsat 8 OLI: 2013-present
rename_l8 <- function(image) {
  image$select(
    c("SR_B2", "SR_B3", "SR_B4", "SR_B5", "SR_B6", "SR_B7", "ST_B10", "QA_PIXEL"),
    c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2", "Thermal", "QA_PIXEL")
  )$multiply(0.0000275)$add(-0.2)$
    copyProperties(image, list("system:time_start", "system:index"))
}
```

## Step 2: Cloud and Shadow Masking

Apply QA_PIXEL bitmask to remove cloud, cloud shadow, and snow pixels.

```{r cloud-mask}
mask_clouds <- function(image) {
  qa <- image$select("QA_PIXEL")

  # Bits 3 (cloud shadow), 4 (cloud), 5 (snow)
  cloud_shadow <- qa$bitwiseAnd(bitwShiftL(1L, 3L))$eq(0L)
  cloud        <- qa$bitwiseAnd(bitwShiftL(1L, 4L))$eq(0L)
  snow         <- qa$bitwiseAnd(bitwShiftL(1L, 5L))$eq(0L)

  mask <- cloud_shadow$And(cloud)$And(snow)

  image$updateMask(mask)$
    select(c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2"))$
    copyProperties(image, list("system:time_start", "system:index"))
}
```

## Step 3: Build the Merged Collection (1994-2015)

Query each sensor for the study period, apply harmonisation and cloud
masking, then merge into a single image collection.

```{r merge-collection}
start_date <- "1994-01-01"
end_date   <- "2016-01-01"

l5 <- ee$ImageCollection("LANDSAT/LT05/C02/T1_L2")$
  filterBounds(aoi_ee)$
  filterDate(start_date, "2012-01-01")$
  filter(ee$Filter$lt("CLOUD_COVER", 30L))$
  map(rename_l5)$map(mask_clouds)

l7 <- ee$ImageCollection("LANDSAT/LE07/C02/T1_L2")$
  filterBounds(aoi_ee)$
  filterDate("1999-01-01", end_date)$
  filter(ee$Filter$lt("CLOUD_COVER", 30L))$
  map(rename_l7)$map(mask_clouds)

l8 <- ee$ImageCollection("LANDSAT/LC08/C02/T1_L2")$
  filterBounds(aoi_ee)$
  filterDate("2013-04-01", end_date)$
  filter(ee$Filter$lt("CLOUD_COVER", 30L))$
  map(rename_l8)$map(mask_clouds)

landsat <- l5$merge(l7)$merge(l8)$sort("system:time_start")

n_scenes <- landsat$size()$getInfo()
print(paste("Total Landsat scenes (1994-2015):", n_scenes))
```

## Step 4: Compute Spectral Water Indices

Five indices from the manuscript, computed on every scene in the merged
collection.

```{r spectral-indices}
add_water_indices <- function(image) {
  blue  <- image$select("Blue")
  green <- image$select("Green")
  red   <- image$select("Red")
  nir   <- image$select("NIR")
  swir1 <- image$select("SWIR1")
  swir2 <- image$select("SWIR2")

  # NDWI: (Green - NIR) / (Green + NIR)
  ndwi <- green$subtract(nir)$divide(green$add(nir))$rename("NDWI")

  # MNDWI: (Green - SWIR1) / (Green + SWIR1)
  mndwi <- green$subtract(swir1)$divide(green$add(swir1))$rename("MNDWI")

  # AWEIsh: Blue + 2.5*Green - 1.5*(NIR + SWIR1) - 0.25*SWIR2
  aweish <- blue$
    add(green$multiply(2.5))$
    subtract(nir$add(swir1)$multiply(1.5))$
    subtract(swir2$multiply(0.25))$
    rename("AWEIsh")

  # WRI: (Green + Red) / (NIR + SWIR1)
  wri <- green$add(red)$divide(nir$add(swir1))$rename("WRI")

  # NDPI: (SWIR1 - Green) / (SWIR1 + Green)
  ndpi <- swir1$subtract(green)$divide(swir1$add(green))$rename("NDPI")

  image$addBands(c(ndwi, mndwi, aweish, wri, ndpi))$
    copyProperties(image, list("system:time_start", "system:index"))
}

landsat_indices <- landsat$map(add_water_indices)
```

## Step 5: Seasonal Composites

Aggregate to seasonal composites (wet: Nov-Apr, dry: May-Oct) per year
using median reduction to suppress outliers.

```{r seasonal-composites}
years <- seq(1994L, 2015L)

make_seasonal <- function(year) {
  # Wet season: November of prior year through April
  wet_start <- ee$Date$fromYMD(as.integer(year - 1L), 11L, 1L)
  wet_end   <- ee$Date$fromYMD(as.integer(year), 5L, 1L)

  # Dry season: May through October
  dry_start <- ee$Date$fromYMD(as.integer(year), 5L, 1L)
  dry_end   <- ee$Date$fromYMD(as.integer(year), 11L, 1L)

  wet <- landsat_indices$filterDate(wet_start, wet_end)$median()$
    set("system:time_start", wet_start$millis())$
    set("year", year)$set("season", "wet")

  dry <- landsat_indices$filterDate(dry_start, dry_end)$median()$
    set("system:time_start", dry_start$millis())$
    set("year", year)$set("season", "dry")

  list(wet, dry)
}

seasonal_list <- unlist(lapply(years, function(y) {
  make_seasonal(y)
}), recursive = FALSE)

landsat_seasonal <- ee$ImageCollection$fromImages(seasonal_list)
print(paste("Seasonal composites:", landsat_seasonal$size()$getInfo()))
```

## Step 6: Spectral Mixture Analysis Endmembers

Define endmember spectra for sub-pixel unmixing. These values should be
refined from training samples, but the initial estimates are drawn from
GEE spectral profiles of known pure pixels within the AOI.

```{r endmembers}
# Sample a clean wet-season composite for endmember extraction
ref_image <- landsat_indices$
  filterDate("2010-01-01", "2010-04-01")$median()$clip(aoi_ee)

# Endmember coordinates (lon, lat) for pure pixels
endmember_points <- list(
  open_water = ee$Geometry$Point(c(35.55, -15.35)),
  flooded_veg = ee$Geometry$Point(c(35.45, -15.25)),
  dry_veg = ee$Geometry$Point(c(35.30, -15.10)),
  bare_soil = ee$Geometry$Point(c(35.65, -15.50)),
  urban = ee$Geometry$Point(c(35.32, -15.39))
)

# Extract spectral signatures
bands <- c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2")

extract_endmember <- function(point, label) {
  values <- ref_image$select(bands)$
    reduceRegion(
      reducer = ee$Reducer$mean(),
      geometry = point,
      scale = 30L
    )$getInfo()
  data.frame(
    endmember = label,
    Blue = values$Blue, Green = values$Green, Red = values$Red,
    NIR = values$NIR, SWIR1 = values$SWIR1, SWIR2 = values$SWIR2,
    stringsAsFactors = FALSE
  )
}

endmembers_df <- do.call(rbind, mapply(
  extract_endmember,
  endmember_points,
  names(endmember_points),
  SIMPLIFY = FALSE
))
print(endmembers_df)
```

## Step 7: Spectral Unmixing

Apply linear spectral unmixing to estimate sub-pixel fractions for each
endmember across the time series.

```{r unmixing}
# Build endmember matrix from extracted values
em_matrix <- as.list(as.data.frame(t(
  endmembers_df[, c("Blue", "Green", "Red", "NIR", "SWIR1", "SWIR2")]
)))

unmix <- function(image) {
  fractions <- image$select(bands)$unmix(em_matrix, TRUE, TRUE)$
    rename(c("water_frac", "flooded_veg_frac", "dry_veg_frac",
             "bare_soil_frac", "urban_frac"))
  image$addBands(fractions)$
    copyProperties(image, list("system:time_start", "system:index"))
}

landsat_unmixed <- landsat_indices$map(unmix)
```

## Step 8: Time Series Extraction

Extract AOI-mean spectral indices and water fraction per scene to build
the analysis time series.

```{r time-series}
extract_ts <- function(image) {
  stats <- image$select(
    c("NDWI", "MNDWI", "AWEIsh", "WRI", "NDPI", "water_frac")
  )$reduceRegion(
    reducer = ee$Reducer$mean(),
    geometry = aoi_ee,
    scale = 30L,
    maxPixels = as.integer(1e9)
  )
  ee$Feature(NULL, stats)$
    set("date", image$date()$format("YYYY-MM-dd"))
}

landsat_ts <- landsat_unmixed$map(extract_ts)

# Pull to R (may take time for large collections)
ts_info <- landsat_ts$getInfo()

landsat_df <- do.call(rbind, lapply(ts_info$features, function(f) {
  p <- f$properties
  data.frame(
    date       = p$date,
    NDWI       = p$NDWI,
    MNDWI      = p$MNDWI,
    AWEIsh     = p$AWEIsh,
    WRI        = p$WRI,
    NDPI       = p$NDPI,
    water_frac = p$water_frac,
    stringsAsFactors = FALSE
  )
}))

landsat_df$date <- as.Date(landsat_df$date)
landsat_df <- landsat_df[order(landsat_df$date), ]
print(paste("Time series records:", nrow(landsat_df)))
```

## Step 9: Quick Plot

Plot the spectral index time series to confirm temporal coverage and
identify recession events.

```{r plot-ts}
ggplot(landsat_df, aes(x = date)) +
  geom_line(aes(y = NDWI, colour = "NDWI"), linewidth = 0.4) +
  geom_line(aes(y = MNDWI, colour = "MNDWI"), linewidth = 0.4) +
  geom_line(aes(y = water_frac, colour = "Water fraction"), linewidth = 0.6) +
  scale_colour_manual(values = c(
    "NDWI" = "steelblue", "MNDWI" = "darkblue", "Water fraction" = "red"
  )) +
  labs(
    x = NULL, y = "Index value / fraction",
    title = "Lake Chilwa spectral water indices (1994-2015)",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Step 10: Export

```{r export}
write.csv(landsat_df,
  "../_staging/landsat_timeseries_indices.csv", row.names = FALSE)

# Export a seasonal composite as GeoTIFF (uncomment as needed):
# task <- ee$batch$Export$image$toDrive(
#   image = landsat_seasonal$
#     filter(ee$Filter$eq("year", 2010L))$
#     filter(ee$Filter$eq("season", "wet"))$
#     first()$
#     select(c("NDWI", "MNDWI", "AWEIsh", "water_frac"))$
#     clip(aoi_ee),
#   description = "Landsat_wet_2010",
#   folder = "GEE_exports",
#   region = aoi_ee,
#   scale = 30L,
#   maxPixels = as.integer(1e9)
# )
# task$start()
```

## Notes

Processing chain: Landsat Collection 2 Level-2 SR from L5-TM, L7-ETM+,
and L8-OLI. Band harmonisation maps sensor-specific band names to a
common schema. QA_PIXEL bitmask removes cloud, shadow, and snow. Five
spectral water indices (NDWI, MNDWI, AWEIsh, WRI, NDPI) computed per
scene. Linear spectral unmixing estimates sub-pixel water fraction from
five endmembers. Seasonal composites (wet: Nov-Apr, dry: May-Oct) built
via median reduction. Scene-level time series extracted as AOI means for
trend analysis.

The scale factor (0.0000275, offset -0.2) converts Collection 2 Level-2
integer values to surface reflectance. Cloud cover threshold is 30%
per scene, consistent with the manuscript QA parameters.
